Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-04-28 _appendix/_model-predictors.qmd
Add appendix for syllable frequencies and new revisions

Our study capitalises in phonology, therefore we operationalised *cognateness* as the phonological similarity between a pair of translation equivalents. Our dataset contained `r max(n_items_final)` translation equivalents, each consisting in two word-forms: Catalan and Spanish (see @tbl-translation-levenshtein for an example). Measuring the phonological similarity between two word forms is non trial. While orthographic similarity can be calculated using string similarity/distance measures on the written word-forms, phonology poses additional challenges. A first, bottom-up approach to to the task might be considering acoustic similarity as a proxy to phonological similarity. This may involve registering audio recordings of a talker reading each word-form aloud, and then comparing their spectrograms,  cochleograms, or formant tracks [e.g., @heeringa2003norwegian]. While this is possible, this method computes acoustic similarity, as opposed to phonological similarity. This is not optimal, since a purely acoustic measure of similarity ignores how the actual signal is perceived by a actual listener. Since approximately 12 months of age, humans perceive the acoustic sounds in speech as phonemes, according to the phonology of their native language(s) [@werker1984crosslanguage; @kuhl1992linguistic]. This means that two listeners may not perceive the same linguistically relevant acoustic signal identically. This is why phonological similarity needs to be computed from discrete units that are meaningful for language perception.

@@

Multiple algorithms have been created to compute the distance between two strings of characters. The Levenshtein distance being one of them. The this algorithm calculates the minimum number of edit operations (additions, deletions, and substitutions) that one string must go through to become identical to the other string [@levenshtein1966binary, see @eq-levenshtein]. For instance, /kaza/ (*house*, in Catalan), would need to go through one substitution to become identical to its translation equivalent in Spanish /kasa/ (/z/ is replaced by /s/). Therefore, the Levenshtein distance between /kaza/ and /kasa/ is one. For an easier handling of non-ASCII characters (fairly prevalent in Catalan and Spanish, e.g., á, ü, ñ), we used phonological transcriptions in X-SAMPA format , as opposed to IPA, for its computer-friendly set of symbols [@wells1995computer].

@@

The R package `stringdist` [@van-der-loo2014stringdist] offers the `stringsim()` function to compute this normalised Levenshtein similarity measure (see @tbl-translation-levenshtein for more examples):

| Catalan 	      | Spanish   	      | Levenshtein  |       
|---------	      |---------	      |-----         |
| *porta* [pOrt5] | *puerta* [pwe4ta] |   0.50 (3)   |
| *taula* [tawl5] | *mesa* [mesa]     |   0.00 (5)   |
| *cotxe* [kOtS5] | *coche* [kotSe]   |   0.40 (3)   |

: Normalised Levenshtein similarity computed for three exemplars of translation pairs in our study {#tbl-translation-levenshtein}

>>>>

Our study capitalises in phonology, therefore we operationalised *cognateness* as the phonological similarity between a pair of translation equivalents. Our dataset contained `r max(n_items_final)` translation equivalents, each consisting in two word-forms: Catalan and Spanish (see {apatb-translation-levenshtein} for an example). Measuring the phonological similarity between two word forms is non trial. While orthographic similarity can be calculated using string similarity/distance measures on the written word-forms, phonology poses additional challenges. A first, bottom-up approach to to the task might be considering acoustic similarity as a proxy to phonological similarity. This may involve registering audio recordings of a talker reading each word-form aloud, and then comparing their spectrograms,  cochleograms, or formant tracks [e.g., @heeringa2003norwegian]. While this is possible, this method computes acoustic similarity, as opposed to phonological similarity. This is not optimal, since a purely acoustic measure of similarity ignores how the actual signal is perceived by a actual listener. Since approximately 12 months of age, humans perceive the acoustic sounds in speech as phonemes, according to the phonology of their native language(s) [@werker1984crosslanguage; @kuhl1992linguistic]. This means that two listeners may not perceive the same linguistically relevant acoustic signal identically. This is why phonological similarity needs to be computed from discrete units that are meaningful for language perception.

@@

Multiple algorithms have been created to compute the distance between two strings of characters. The Levenshtein distance being one of them. The this algorithm calculates the minimum number of edit operations (additions, deletions, and substitutions) that one string must go through to become identical to the other string [@levenshtein1966binary] (see @eq-levenshtein). For instance, /kaza/ (*house*, in Catalan), would need to go through one substitution to become identical to its translation equivalent in Spanish /kasa/ (/z/ is replaced by /s/). Therefore, the Levenshtein distance between /kaza/ and /kasa/ is one. For an easier handling of non-ASCII characters (fairly prevalent in Catalan and Spanish, e.g., á, ü, ñ), we used phonological transcriptions in X-SAMPA format , as opposed to IPA, for its computer-friendly set of symbols [@wells1995computer].

@@

The R package `stringdist` [@van-der-loo2014stringdist] offers the `stringsim()` function to compute this normalised Levenshtein similarity measure, see {apatb-translation-levenshtein} for more examples:

```{r apatb-translation-levenshtein}
#| label: apatb-translation-levenshtein
#| apa-cap: Normalised Levenshtein similarity computed for three exemplars of translation pairs in our study
tribble(~Catalan, ~Spanish, ~Levenshtein,
        "porta [pOrt5]", "puerta [pwe4ta]", 0.50,
        "taula [tawl5]", "mesa [mesa]", 0.00,
        "cotxe [kOtS5]", "coche [kotSe]", 0.40) |> 
    gt() |> 
    fmt_number(is.numeric)
```
